---
title: "geseca-tutorial"
# output: rmarkdown::html_vignette
output:
  md_document:
    variant: markdown_github
vignette: >
  %\VignetteIndexEntry{geseca-tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

An R-package for performing GEne SEt Co-regulation Analysis (GESECA).

## Quick run

Loading necessary libraries

```{r message=FALSE}
library(fgsea)
library(data.table)
```

```{r echo=FALSE}
library(BiocParallel)
register(SerialParam())
```

Loading example expression matrix and pathways:

```{r message=FALSE}
data(exampleExpressionMatrix)
str(exampleExpressionMatrix)
```
`exampleExpressionMatrix` - numeric expression matrix of the GSE14308 dataset. Rows correspond to genes (ENTREZID is used as identifiers), columns correspond to samples. **Note:** `geseca` expects that the values of expression matrix are log-scaled and the samples are quantile normalized. To get more information about steps that were performed to process the GSE14308 expression matrix, use: `?exampleExpressionMatrix`.


```{r message=FALSE}
data(examplePathways)
str(head(examplePathways, 3))
```
`examplePathways` - list of mouse pathways from `reactome.db` package. The list of gene sets should use the same identifiers that are used in `rownames(exampleExpressionMatrix)`. To get more information about this list, use: `?examplePathways`

Running `geseca`:
```{r}
gesecaRes <- geseca(exampleExpressionMatrix, examplePathways, minSize = 15, maxSize = 500)
head(gesecaRes[order(pval), ], 5)
```
Optionally, it is possible to scale the values of each gene-row to a unit variance by passing the `scale=TRUE` to `geseca`.
